\chapter{5417E Spreadsheet Hardware Integrated Technology Engine}
\label{cha:shite}

\section{Overview}

Introducing the MEGA SheetStorm 65 Spreadsheet Hardware Integrated
Technology Engine (aka SHITE), the groundbreaking chip designed
for the MEGA65 system, set to revolutionize spreadsheet processing in
ways the industry has never seen before. With the innovative "Data
Fan" technology, this chip is not just an improvement over existing
solutions; it's a complete redefinition of speed and efficiency in
data management. 

When your sheet data hits the Data Fan of the MEGA SheetStorm 65, it's
an experience unlike any other. The Data Fan, a unique innovation in
data distribution, ensures that your cells are coated with updated
data faster and more thoroughly than you've ever witnessed. It's like
a gust of fresh air sweeping through your spreadsheets, revitalizing
every cell with newfound speed and accuracy. In the world of data
processing, if speed is the game, the MEGA SheetStorm 65 is the
undisputed champion. 

The power of the MEGA SheetStorm 65 is so formidable that it
effectively flushes away the competition. Where others falter and clog
under the pressure of massive data sets, the SheetStorm 65 maintains a
flow of information so smooth, it’s like watching a masterfully
engineered system handle what others would consider a deluge of
data. This is not just processing; this is data management artistry. 

In today's fast-paced business world, being bogged down by sluggish
data processing is like throwing money down the drain. The MEGA
SheetStorm 65 addresses this head-on, offering a level of efficiency
that not only keeps your data flowing but also enhances the overall
health of your information systems. It's a chip that doesn’t just work
hard; it works smart, ensuring that every bit of data is where it
needs to be, exactly when it needs to be there. 

In summary, the MEGA SheetStorm 65 is more than just a technological
advancement; it’s a revolution in spreadsheet processing. It's an
affirmation that in the realm of data, the right tool doesn't just
make a difference; it makes all the difference. With the SheetStorm
65, prepare to see your data, and your business, in a whole new
light. Welcome to the future of data processing. Welcome to the MEGA
SheetStorm 65.

\section{Architecture}

The SHITE processor consists of a shared memory mapped region that
allows both the main processor and the SHITE processor simultaneous
access to spread-sheet data.  Correctly described spreadsheet cells
are continously updated at high-speed in hardware by the SHITE, providing a
dramatically faster performance than can be obtained using a
convention processor.

For rapid access by the processor, a sub-set of these cells can also
be plumbed into 16-bit address space using the Streamlined Electronic
Workflow for Easy Retrieval (SEWER) connection.

\section{Unimplemented Functionality}
The SHITE processor is in the process of being implemented, and is not
currently available. It is expected to be made available in a future
core update of the MEGA65.
