\chapter{Introduction}


\section{Welcome to the MEGA65!}

Congratulations on your purchase of one of the most long-awaited computers in the history of computing! This computer is community designed, and based on the never-released Commodore{\textregistered} 65\footnote{Commodore is a trademark of C= Holdings} computer; a computer designed in 1989 and intended for public release in 1990. Decades have passed, and we have endeavoured to invoke an earlier time when computers were simple and friendly. They were not only simple to operate and understand how they work, but friendly and approachable for new users.

These 1980s computers inspired an entire generation of professionals to choose the exciting and rewarding technology careers they have today. Just imagine the joy of these individuals as they learned they could use their new computer to solve problems, write a letter, prepare taxes, invent new things, or even discover how the universe works. We want to re-create that level of excitement not found in modern computing, so we created the {\bf MEGA65}.

The MEGA65 team believes that owning a computer is like owning a home. You don't just use a home; you change things, big and small, to make it your own custom living space. After a while, when you settle in, you may decide to renovate or expand your home to make it more comfortable, or provide more utility. Think of the MEGA65 as a ``computing home''.

This guide will teach you how to do more than just hang pictures on a wall; it will show you how to build your dream home. While you read this user's guide, you will learn how to operate the MEGA65, write programs, add additional software, and extend hardware capabilities. What won't be immediately obvious is that along the journey, you will also learn about the history of computing as you explore BASIC version 65 and operating system commands.

Computer graphics and music make computing more fun, and we designed the MEGA65 to be fun! In this user's guide, you will learn how to write programs using the MEGA65's built-in {\bf graphics} and {\bf sound} capabilities. But you don't need to be a programmer to have fun with the MEGA65. Because the MEGA65 includes a complete Commodore{\textregistered} 64{\texttrademark}\footnote{Commodore 64 is a trademark of C= Holdings}, it can also run thousands of existing games, utilities, and business software packages, as well as new programs being written today by Commodore computer enthusiasts. Excitement for the MEGA65 will grow as we discover what programmers create, as they discover and master the powerful capabilities of this modern Commodore computer recreation. Together, we can build a new ``homebrew'' community, creating software and projects that push the MEGA65's capabilities far beyond what anyone thought would be possible.

We welcome you on this journey! Thank you for becoming a part of the MEGA65
community of users, programmers, and enthusiasts!

\section{Other Books in this series}

This is one of several MEGA65 documentation volumes. The series includes:

\begin{itemize}
    \item The MEGA65 User's Guide -- Provides an introduction to the MEGA65, and a condensed BASIC 65 command reference
    \item The MEGA65 BASIC 65 Reference Guide -- Comprehensive documentation of all BASIC 65 commands, functions and operators
    \item The MEGA65 Chipset Reference -- Detailed documentation about the MEGA65 and C65's custom chips
    \item The MEGA65 Developer Guide -- Information for developers who wish to write programs for the MEGA65
    \item The MEGA65 Book -- All volumes in a single huge PDF for easy searching. 1080 pages and growing!
\end{itemize}

\section{Come Join Us!}
Get involved, learn more about your MEGA65, and join us online at:

\begin{itemize}
    \item \url{http://mega65.org/chat}
    \item \url{http://mega65.net}
\end{itemize}
